Measuring directionality in double-beta decay and neutrino interactions with kiloton-scale scintillation detectors

### 2014-01-08 Direction_JINST_v7.tex
Added changes from Henry from the weekend. --Lidnley

Scintillation light is isotropic. At these low energies, it does not contain sufficient to reconstruct the track of the outgoing particles, although at higher energies it may\cite{john}.  Cherenkov light is also
produced for electrons above threshold. Most is absorbed and re-emitted as part of the
scintillation processes; however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection, especially for $0\nu\beta\beta$ searches. In the high pressure TPCs, reduction factors on the order of $10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches and merits investigation.

This is the first in a series of papers exploring directionality in large-scale liquid scintillator detectors. In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector advances in timing, photodetector 
spectral response, and scintillator emission spectra that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors. This is a different technique and application from the direction reconstruction described for high-energy neutrino interactions\cite{john} or that for neutrons from inverse beta decay\cite{chooz,dcDirection}.  We then use these results as input into a traditional direction reconstruction developed for water Cherenkov detectors. Since the reconstruction of the direction of $\sim$1~MeV particles has not been achieved before, we start these studies with the most simple case of a single particle at the center of the detector. We also start with an easier test case of a 5~MeV electron like that from a $^{8}$B solar neutrino interaction. With the higher photon statistics at this energy, we verify the technique and the different detector parameters that affect it. We then lower the energy to that relevant to a single electron in a $0\nu\beta\beta$.

@@

The Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. For the example scintillator above, n(400nm)=1.459\cite{OlegThesis}. The absorption cutoff is between 370~nm and 360~nm where the attenuation length rapidly drops from 6.5~m to 0.65~m. For a 5~MeV electron, this yields 744 Cherenkov photons per event; dropping down to a 1~MeV electron yields 89 Cherenkov photons per event when integrating from the cutoff wavelength at 360~nm to 550~nm, the point above which a small number of photons is detected due to a standard photocathode's quantum efficiency. For reference at this cutoff wavelength, the index of refraction n(370nm)=1.466. This can be translated into an effective Cherenkov threshold for electrons at 0.188~MeV.

All photons including these undisturbed Cherenkov
photons will have timing determined by the group 
velocity\cite{group_velocity_article,pdg_review_2012,tamm1939} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
This means photons at the scintillation cutoff at 370~nm will travel at 0.191~m/ns while those at 600~nm travel at 0.203 m/ns. Since on average the undisturbed Cherenkov photons have longer wavelengths, they will arrive before the
scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations which depend on the underlying $0\nu\beta\beta$ decay mechanism\cite{SuperNEMO,newphysics0nuBB,bandv}.
For light Majorana neutrino exchange, a probable case is the equal division of energy between two
electrons emitted at a large angle relative to each other\cite{phasespace, SuperNEMO}. An example case is shown in figure \ref{detector_view}
for a $^{116}$Cd $0\nu\beta\beta$ event. Since the decay
half-life is inversely proportional to the phase-space factor, isotopes with
higher Q-values are preferred. Due to backgrounds from the daughters
of the $^{238}$U and $^{232}$Th decay chains, isotopes with Q-values above the 2.6~MeV gamma from $^{208}$Tl are preferred; however slightly lower Q-values can be tolerated if there is sufficient background rejection. There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb},
but only a handful with sufficiently large Q-values.

Most of the high Q-value candidates
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd (Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr (Q=3.350~MeV)\cite{zr1},
$^{100}$Mo (Q=3.034~MeV)\cite{mo1}, $^{82}$Se (Q=2.995~MeV)\cite{qdot},
$^{116}$Cd (Q=2.81~MeV)\cite{qdot, cd1}, $^{130}$Te (Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe (Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn (Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
a suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. Recently nanocrystals formed by candidate isotopes have been explored as an alternative to doping by single atoms,\cite{qdot,qdot2}.

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs2/Cerenkov_PEs_scattering_comparison.pdf}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV). The default simulation (black) is compared to the simulation with optical scattering turned on (red). Scattering causes a small tail to longer times as expected. \label{scattplot}}
        \end{center}
\end{figure}

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. Similar to optical scattering, reflections at the sphere are a small effect that would create a small tail at longer times. Two important photodetector properties have been varied: 1)
the transit-time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}. The QE values as a function of wavelength come from the Double Chooz\cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency.

\begin{figure}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut, 171~PEs
        from scintillation and 108~PEs from Cherenkov light are detected. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut, 349~PEs from
        scintillation and 88~PEs from Cherenkov light are detected. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut, 226~PEs from scintillation and 229~EPS
        from Cherenkov light are detected. \label{time_plots_comparison}}
        \end{center}
\end{figure}

@@

Four effects primarily contribute to the timing of the scintillator detector
system: the travel time of the particle, the time constants of the scintillation process, chromatic dispersion and the timing of the photodetector.  First, the simulated travel time of a 5~MeV electron is 0.108$\pm$0.015~ns. This corresponds to an average path length of 3.1~cm and a final distance from the origin of 2.6~cm. The time until the electron drops below Cherenkov threshold is 0.106$\pm$0.015~ns. We note that due to scattering the final direction of the electron before it stops does not correspond to the initial direction; however before the electron drops below Cherenkov threshold the scattering angle is small and therefore the Cherenkov light encodes the direction of the primary electron. The scattering physics is handled by Geant4's ``Multiple Scattering" process which is valid down to 1~keV, where atomic shell structure becomes important\cite{geant4scatt}. In the energy range important for $0\nu\beta\beta$, a 1.4~MeV electron travels a total path length of 0.8~cm, has a distance from the origin of 0.6~cm in 0.030$\pm$0.004~ns  and takes 0.028$\pm$0.004~ns to drop below Cherenkov threshold. The scattering follows the same pattern.

@@

Due to the wavelength-dependence of the refractive index the speed of
light in the scintillator (see Equation (\ref{eqGroup})) increases
with increasing photon wavelengths for normal dispersion, with red
light traveling faster than blue light.  In order to study the
measurability of the time differences due to this chromatic dispersion, we simulated 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including a
transit-time spread. The higher energy provides larger photon statistics with which to evaluate the method, debug the simulation and is of interest to neutrino-electron scattering experiments. The true hit time distributions of photoelectrons
were analyzed for scintillation light and Cherenkov light
separately. Photoelectrons coming from Cherenkov light are on average
created about 0.5~ns earlier than PEs from scintillation light. The
RMS values from PE time distributions for Cherenkov and scintillation
light are both about 0.5~ns. Note that these numbers include the
effect of the finite electron travel time.

@@

In Sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations at 5~MeV. We focus
on the idea of increasing the discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the \GEANT~simulation are the photoelectron hit
positions and the detection times after the TTS resolution has been
applied. In Section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction. With the successful reconstruction at 5~MeV, we then lower the energy of the simulated electrons and show that it is possible to reconstruct electrons in the range interesting for $0\nu\beta\beta$.

@@

In order to compare simulations with different parameters, a fixed
time cut of $t\leq$ 34.0~ns is applied using the truth information to
isolate the Cherenkov light in this early time window, see figure \ref{time_plots_comparison} (a). For the default
simulation case, the average number of PEs per event coming from
Cherenkov light in the early time window (108) is 98\% of the total
average number of PEs from Cherenkov light (110). For scintillation
light, the average number of PEs (171) is only 3.1\% of the average total
scintillation-induced PEs (5445). This demonstrates the effectiveness of
a time cut to separate Cherenkov light from scintillation light.

@@

When the 17-inch KamLAND PMTs\cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cherenkov over
scintillation light ($R_{C/S}=0.25$) for $t<34.0$~ns (see
figure \ref{time_plots_comparison} (b)). This shows that a low
photodetector TTS is critical for directionality reconstruction and
motivates the use of novel photodetector types.

\section{Detector wavelength response}
\label{detector_wavelength_response_sec} 
In addition to decreasing the photodetector TTS to enhance $R_{C/S}$,
it is possible to optimize the wavelength-dependence of the
photocathode. Since Cherenkov photons that pass through meters of
scintillator have on average longer wavelengths than scintillation
photons, a photodetector that is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs.

We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63)\cite{Hamamatsu_R3899U}.
Figure \ref{time_plots_comparison}(c) shows the results for the
modified simulation with high QE in the red spectral region. The
higher absolute number of photoelectrons coming from Cherenkov light
(factor of $\approx$ 2) and the increased Cherenkov/scintillation
ratio ($R_{C/S}=1.01$) in the early time window would significantly improve the
directionality reconstruction.

@@

The timing studies show that in the early time window, $t\leq$
34.0~ns, the ratio $R_{C/S}$ is high, improving the photoelectron hit selection. In this section, we apply
reconstruction tools for a water Cherenkov detector, WCSimAnalysis,
to the problem of reconstructing the position and direction of 5~MeV
electrons from this early light. To start, we are neglecting the effects of position dependence. In future work the arrival times can be corrected by the time of flight from the reconstructed vertex and the position and direction fitted simultaneously. The vertex reconstruction uncertainty leads to an additional smearing of the arrival time distribution of 0.15~ns for every 3~cm of position reconstruction uncertainty.

WCSimAnalysis is a water Cherenkov
reconstruction package developed by the Long Baseline Neutrino
Experiment (LBNE) Collaboration\cite{Blake}. It provides a framework
for generic event cleaning, track reconstruction, and particle
identification, and comes equipped with variety of pre-built
algorithms. It is continuing to be expanded using new track-fitting
techniques for water Cherenkov detectors\cite{Sanchez2012525} based on
advanced photosensors with sub-cm imaging capabilities and timing
resolutions below 100 picoseconds.

The results presented in this paper rely on a simple vertex
reconstruction algorithm, commonly known as a ``point
fit''\cite{SuperKalgo}. It assumes that all of the scintillation and
Cherenkov light is emitted from a single point in space-time
$(x_0,y_0,z_0,t_0)$. In actuality, the light is emitted along a multi-scattered electron track. However, at the energies
discussed in this paper, the extent of this track is small (a few cm)
compared to the scale of the detector (R=6.5 meters), which sets the typical
photon transit distances.

@@

Once a set of vertex candidates has been found, we test the goodness
of each vertex and select the one that best fits the full ensemble of
photon hits. The goodness of fit is determined based on the
distribution of an observable known as the ``point time
residual''\cite{SuperKalgo}. The point time residual is calculated by
taking the difference between the measured time of a photon hit, and
the predicted time of the hit, given its distance from the vertex
hypothesis, a single effective speed of light in the scintillator, and
the hypothesized $t_0$ of the event. The width of the time residual
distribution over all hits is minimized when the hypothesized vertex
is near the true vertex. Based on this figure of merit, we select the
vertex with the narrowest time residual distribution from among the
400 candidates.

@@

For the purpose of testing the reconstruction algorithm we use 1000
simulated electrons with an energy of 5~MeV, lower energies are studied in the next section.
The electrons are simulated at the center of the detector, $\vec{r}$ = (0,0,0), along
the x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0). Figure \ref{fig:reco} (Top)
shows the vertex reconstruction. The vertex is reasonably well
reconstructed around the center of the detector, $\vec{r}$ = (0,0,0),
except along the x-axis. The RMS values of the distributions for all three
reconstructed coordinates are smaller than 3.5~cm. The shift along
the x-axis is due to two effects for which the reconstruction has to
use average values rather than the unknown true value for each
hit: the wavelength and hence the speed of the light in the medium,
and the point of emission for each of the photons, reconstructed
as coming from a common point. The reconstruction of the direction
also is shown in figure \ref{fig:reco} (Top). It shows that for the majority
of the events the initial electron direction is reconstructed well.
This is a promising result given the simplicity of the algorithms
used.

@@

The reconstruction algorithms outlined in section \ref{reconstruction_sec} have also been applied
to the simulations at lower energies. Figure \ref{fig:reco} (Middle) shows the results for 2.1~MeV and Figure \ref{fig:reco} (Bottom) shows the results for 1.4~MeV. Most events are still reconstructed well, despite the lower
number of PEs and the decreased $R_{C/S}$. For 1.4~MeV electrons, the RMS values of the distributions for all three
reconstructed coordinates are smaller than 4.5~cm. The mean cosine of the angle between the true direction
and the reconstructed direction for different energies of the initial electron is shown in figure
\ref{Edep_angle}. The direction reconstruction performance is still promising for energies as low as 1.4~MeV.

>>>>

Scintillation light is isotropic. At these low energies, it does not contain sufficient information to reconstruct the track of the outgoing particles, although at higher energies it may\cite{john}.  Cherenkov light is also
produced for electrons above threshold. Most is absorbed and re-emitted as part of the
scintillation processes; however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection, especially for $0\nu\beta\beta$ searches. In the high pressure TPCs, reduction factors on the order of $10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches and merits investigation.

This is the first in a series of papers exploring directionality in large-scale liquid scintillator detectors. In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector advances in timing, photodetector 
spectral response, and scintillator emission spectra that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors. This is a different from the direction reconstruction described for high-energy neutrino interactions\cite{john} or that for neutrons from inverse beta decay\cite{chooz,dcDirection}.  We then use these results as input into a traditional direction reconstruction developed for water Cherenkov detectors. Since the reconstruction of the direction of $\sim$1~MeV particles has not been achieved before, we start these studies with the most simple case of a single particle at the center of the detector. We also start with an easier test case of a 5~MeV electron like that from a $^{8}$B solar neutrino interaction. With the higher photon statistics at this energy, we verify the technique and the different detector parameters that affect it. We then study the technique for two lower energies, 1.4 and 2.1 MeV, that are more relevant to $0\nu\beta\beta$.

@@

The Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. The absorption cutoff for the example scintillator above is 370~nm. The index of refraction at  n(370nm)=1.466 and can be translated into an effective Cherenkov threshold for electrons at 0.188~MeV. For a 5~MeV electron, this yields 685 Cherenkov photons per event when integrating from the cutoff wavelength at 360~nm to 550~nm,  the wavelength above which a small number of photons is detected due to a standard bialkali photocathode's quantum efficiency. Lowering the energy to 1~MeV yields 82 Cherenkov photons between 360-550~nm.

%For the example scintillator above, n(400nm)=1.459\cite{OlegThesis}. The absorption cutoff is between 370~nm and 360~nm where the attenuation length rapidly drops from 6.5~m to 0.65~m. For a 5~MeV electron, this yields 744 Cherenkov photons per event; dropping down to a 1~MeV electron yields 89 Cherenkov photons per event when integrating from the cutoff wavelength at 360~nm to 550~nm, the point above which a small number of photons is detected due to a standard photocathode's quantum efficiency. For reference at this cutoff wavelength, the index of refraction n(370nm)=1.466. This can be translated into an effective Cherenkov threshold for electrons at 0.188~MeV.

All photons including these undisturbed Cherenkov
photons will have timing determined by the group 
velocity\cite{group_velocity_article,pdg_review_2012,tamm1939} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
Photons at the scintillation cutoff of 370nm have a velocity of 0.191 m/ns, while photons with wavelengths of 600 nm are appreciably faster, with a velocity of 0.203 m/ns. Since on average the undisturbed Cherenkov photons have longer wavelengths, they will arrive before the
scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

In neutrino-electron scattering events, a single electron emerges with a distribution of energies with a maximum energy related to the incoming neutrino's energy. In comparison, $0\nu\beta\beta$ events are more complicated with two electron emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations which depend on the underlying $0\nu\beta\beta$ decay mechanism\cite{SuperNEMO,newphysics0nuBB,bandv}. Figure 1 shows a simulated $^{116}$Cd $0\nu\beta\beta$ event in a model with light Majorana neutrino exchange, for which a probable case is the emission of two electrons with comparable energies at a large angle relative to each other.

High Q-value candidates are preferred for $0\nu\beta\beta$  for two reasons. First, the measured half-life is inversely proportional to the phase-space factor, therefore the measured rate is expected to be higher from isotopes with higher Q-values. Second, the main background for these experiments come from the daughters of the $^{238}$U and $^{232}$Th decay chains. The Compton shoulder from gamma-rays is particularly problematic. The 2.6~MeV gamma-ray from $^{208}$Tl decay is the highest energy gamma from these decay chains and therefore isotopes above 2.6~MeV are preferred.  Experiments using isotopes with Q-values below this energy must compensate with improved background rejection techniques.

Most of the high Q-value candidates\cite{tabledbb}
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd (Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr (Q=3.350~MeV)\cite{zr1},
$^{100}$Mo (Q=3.034~MeV)\cite{mo1}, $^{82}$Se (Q=2.995~MeV)\cite{qdot},
$^{116}$Cd (Q=2.81~MeV)\cite{qdot, cd1}, $^{130}$Te (Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe (Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn (Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
a suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. Recently nanocrystals formed by candidate isotopes have been explored as an alternative to doping by single atoms,\cite{qdot,qdot2}.

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs2/PEtime_Cerenkov_scattered_vs_unscattered.pdf}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV). The default simulation is shown with optical scattering turned on: all photons (black), un-scattered photons (red dotted), and scattered (blue dashed). Scattering causes a very small tail at longer times as expected. \label{scattplot}}
        \end{center}
\end{figure}

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. As in the case of optical scattering, reflections at the sphere are a small effect that would create a small tail at longer times. Two important photodetector properties have been varied: 1)
the transit-time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}. The QE values as a function of wavelength come from the Double Chooz\cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency.

\begin{figure}
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut, 171~PEs
        from scintillation and 108~PEs from Cherenkov light are detected. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17-inch PMTs). After the 34.0~ns time cut, 349~PEs from
        scintillation and 88~PEs from Cherenkov light are detected. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut, 226~PEs from scintillation and 229~EPS
        from Cherenkov light are detected. \label{time_plots_comparison}}
        \end{center}
\end{figure}

@@

Four effects primarily contribute to the timing of the scintillator detector
system: the travel time of the particle, the time constants of the scintillation process, chromatic dispersion, and the timing of the photodetector.  First, the simulated travel time of a 5~MeV electron is 0.108$\pm$0.015~ns. This corresponds to an average path length of 3.1~cm and a final distance from the origin of 2.6~cm. The time until the electron drops below Cherenkov threshold is 0.106$\pm$0.015~ns. We note that due to scattering the final direction of the electron before it stops does not correspond to the initial direction; however before the electron drops below Cherenkov threshold the scattering angle is small and therefore the Cherenkov light encodes the direction of the primary electron. The scattering physics is handled by Geant4's ``Multiple Scattering" process which is valid down to 1~keV, where atomic shell structure becomes important\cite{geant4scatt}. In the energy range important for $0\nu\beta\beta$, a 1.4~MeV electron travels a total path length of 0.8~cm, has a distance from the origin of 0.6~cm in 0.030$\pm$0.004~ns  and takes 0.028$\pm$0.004~ns to drop below Cherenkov threshold. The scattering follows the same pattern.

@@

Due to the wavelength-dependence of the refractive index the speed of
light in the scintillator (see Equation (\ref{eqGroup})) increases
with increasing photon wavelengths for normal dispersion, with red
light traveling faster than blue light.  In order to study the
measurability of the time differences due to this chromatic dispersion, we simulated 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including a
transit-time spread. The higher energy provides larger photon statistics with which to evaluate the method, debug the simulation, and is of interest to neutrino-electron scattering experiments. The true hit time distributions of photoelectrons
were analyzed for scintillation light and Cherenkov light
separately. Photoelectrons coming from Cherenkov light are on average
created about 0.5~ns earlier than PEs from scintillation light. The
RMS values from PE time distributions for Cherenkov and scintillation
light are both about 0.5~ns. Note that these numbers include the
effect of the finite electron travel time.

@@

In sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations at 5~MeV. We focus
on the idea of increasing the discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the \GEANT~simulation are the photoelectron hit
positions and the detection times after the TTS resolution has been
applied. In section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction. With the successful reconstruction at 5~MeV, we then lower the energy of the simulated electrons and show that it is possible to reconstruct electrons in the range interesting for $0\nu\beta\beta$.

@@

In order to compare simulations with different parameters, a fixed
time cut of $t\leq$ 34.0~ns is applied using the truth information to
isolate the Cherenkov light in this early time window, as shown in figure \ref{time_plots_comparison} (a). For the default
simulation case, the average number of PEs per event coming from
Cherenkov light in the early time window (108) is 98\% of the total
average number of PEs from Cherenkov light (110). For scintillation
light, the average number of PEs (171) is only 3.1\% of the average total
scintillation-induced PEs (5445). This demonstrates the effectiveness of
a time cut to separate Cherenkov light from scintillation light.

@@

When the 17-inch KamLAND PMTs\cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cherenkov over
scintillation light ($R_{C/S}=0.25$) for $t<34.0$~ns (see
figure \ref{time_plots_comparison} (b)). This shows that a
photodetector with a low TTS is critical for directionality reconstruction and
motivates the use of novel photodetector types.

\section{Detector wavelength response}
\label{detector_wavelength_response_sec} 
Since Cherenkov photons that pass through meters of
scintillator have on average longer wavelengths than scintillation
photons, a photodetector that is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs. We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63)\cite{Hamamatsu_R3899U}, but with the default TTS of 0.1~ns.
Figure \ref{time_plots_comparison}(c) shows the results for the
modified simulation with high QE in the red spectral region. The
higher absolute number of photoelectrons coming from Cherenkov light
(factor of $\approx$ 2) and the increased Cherenkov/scintillation
ratio ($R_{C/S}=1.01$) in the early time window would significantly improve the
directionality reconstruction.

@@

The timing studies show that in the early time window, $t\leq$
34.0~ns, the ratio $R_{C/S}$ is high, improving the photoelectron hit selection. In this section, we apply
reconstruction tools for a water Cherenkov detector, WCSimAnalysis,
to the problem of reconstructing the position and direction of 5~MeV
electrons from this early light. WCSimAnalysis is a water Cherenkov
reconstruction package developed by the Long Baseline Neutrino
Experiment (LBNE) Collaboration\cite{Blake}. It provides a framework
for generic event cleaning, track reconstruction, and particle
identification, and comes equipped with variety of pre-built
algorithms. It is continuing to be expanded using new track-fitting
techniques for water Cherenkov detectors\cite{Sanchez2012525} based on
advanced photosensors with sub-cm imaging capabilities and timing
resolutions below 100 picoseconds.

To start, we are neglecting the effects of position dependence. In future work the arrival times can be corrected by the time of flight from the reconstructed vertex and the position and direction fitted simultaneously. For isotropic light, the vertex reconstruction uncertainty leads to an additional smearing of the arrival time distribution of 0.15~ns for every 3~cm of position reconstruction uncertainty. For directional light, the vertex reconstruction uncertainty leads to an effective shift of the arrival time distribution. This would change the ideal time cut by 0.15~ns  for every 3~cm of position reconstruction uncertainty. We have studied other time cuts from 33~ns to 34.5~ns and the reconstruction remained reasonable.

The results presented in this paper rely on a simple vertex
reconstruction algorithm, commonly known as a ``point
fit''\cite{SuperKalgo}. It assumes that all of the scintillation and
Cherenkov light is emitted from a single point in space-time
$(x_0,y_0,z_0,t_0)$. In actuality, the light is emitted along a multi-scattered electron track. However, at the energies
discussed in this paper, the extent of this track is small (a few cm)
compared to the scale of the detector (R=6.5 meters) that sets the typical
photon transit distances.

@@

Once a set of vertex candidates has been found, we test the goodness
of each vertex and select the one that best fits the full ensemble of
photon hits. The goodness of fit is determined based on the
distribution of an observable known as the ``point time
residual''\cite{SuperKalgo}. The point time residual is calculated by first choosing a hypothesis for the vertex position and $t_0$ for the event. The goodness of fit for these values  is then calculated by taking the difference between the measured and predicted times of each photon hit, using a single effective speed of light in the scintillator. The width of the time residual
distribution over all hits is minimized when the hypothesized vertex
is near the true vertex. Based on this figure of merit, we select the
vertex with the narrowest time residual distribution from among the
400 candidates.

@@

For the purpose of testing the reconstruction algorithm we use 1000
simulated electrons with an energy of 5~MeV; lower energies are studied in the next section.
The electrons are simulated at the center of the detector, $\vec{r}$ = (0,0,0), along
the x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0). Figure \ref{fig:reco} (Top)
shows the vertex reconstruction. The vertex is reasonably well-reconstructed around the center of the detector, $\vec{r}$ = (0,0,0),
except along the x-axis. The RMS values of the distributions for all three
reconstructed coordinates are smaller than 3.5~cm. The shift along
the x-axis is due to two effects for which the reconstruction has to
use average values rather than the unknown true value for each
hit: the wavelength and hence the speed of the light in the medium,
and the point of emission for each of the photons, reconstructed
as coming from a common point. The reconstruction of the direction
also is shown in figure \ref{fig:reco} (Top). It shows that for the majority
of the events the initial electron direction is reconstructed well.
This is a promising result given the simplicity of the algorithms
used.

@@

The reconstruction algorithms outlined in section \ref{reconstruction_sec} have also been applied
to the simulations at lower energies. Figure \ref{fig:reco} (Middle) shows the results for 2.1~MeV and figure \ref{fig:reco} (Bottom) shows the results for 1.4~MeV. Most events are still reconstructed well, despite the lower
number of PEs and the decreased $R_{C/S}$. For 1.4~MeV electrons, the RMS values of the distributions for all three
reconstructed coordinates are smaller than 4.5~cm. The mean cosine of the angle between the true direction
and the reconstructed direction for different energies of the initial electron is shown in figure
\ref{Edep_angle}. The direction reconstruction performance is still promising for energies as low as 1.4~MeV.
